---
title: "Trees in Paris"
author: "Eric Marcon & Florence Puech"
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
bibliography: entropie.bib
output:
  ioslides_presentation:
    logo: images/EcoFoG2017.png
    widescreen: true
  beamer_presentation:
    df_print: kable
    fig_caption: no
    includes:
      in_header: EcoFoGBeamer.tex
    keep_tex: no
    slide_level: 2
  slidy_presentation: default
---

```{r setup, include=FALSE}
### knitr and R options (packages, ...)
# echo=FALSE not to display R chunk in slides. Set to TRUE if necessary.
knitr::opts_chunk$set(echo = FALSE, cache=TRUE)
# tidy R chunks.
knitr::opts_chunk$set(tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=50))
# R console width
options(width=50)
CRANLibrary <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package, repos="https://cran.rstudio.com/")}
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Packages sur GitHub
GitHubLibrary <- function(Packages) {
  InstallAndLoad <- function(Package) {
    Package_split <- stringr::str_split(Package, "/", simplify = TRUE)
    if (!Package_split[1, 2] %in% installed.packages()[, 1]) {remotes::install_github(Package)}
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
CRANLibrary(c("alphahull", "entropart", "kableExtra", "rgdal", "tidyverse"))
GitHubLibrary("EricMarcon/SpatDiv")
# Set ggplotting to bw plot default, but with transparent background elements.  
theme_set(theme_bw(base_size=12))
theme_update(panel.background = element_rect(fill = "transparent", colour = NA),
             plot.background = element_rect(fill = "transparent", colour = NA))
knitr::opts_chunk$set(dev.args=list(bg="transparent"))
# Tibbles: 5 lines, fit to slide width
options(tibble.print_min = 5, tibble.width = 50)
```


# Motivation

<!-- Code HTML complémentaire après le titre de première partie pour éviter une diapo blache --> 
<!-- Commande de saut de colonne HTML --> 
<style>
  .forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>


## Methodological exercise

Recent developments in biodiversity measurement.

Possible interfertilisation with measures of specialisation and spatial concentration in economics.

Applied to Paris trees data.


## Aims

Show similar approaches in different disciplines.

Discuss economic issues:

- MAUP

- absolute vs. relative concentration

Illustrate with trees rather than shops or plants.


## Data

Paris open data: https://opendata.paris.fr

Dataset: 200,000 trees in Paris located in 20 arrondissements (areas), excluding Bois de Vincenne and Bois de Boulogne.

Missing data: Jardin du Luxembourg, owned by Senate and a few others.


```{r data, include=FALSE}
# Trees
if (!file.exists("data/les-arbres.geojson"))
  # Download the file if necessary (100 Mo)
  download.file("https://opendata.paris.fr/explore/dataset/les-arbres/download/?format=geojson", destfile="data/les-arbres.geojson")

# Arrondissements: a polygon layer with administrative limits
if (!file.exists("data/arrondissements.geojson"))
  download.file("https://opendata.arcgis.com/datasets/8463ae61c7ba4388bf08189effcbf2f1_0.geojson", destfile="data/arrondissements.geojson")

library("rgdal")
# Read the GeoJSon objects
les_arbres <- readOGR("data/les-arbres.geojson","les-arbres", encoding = "UTF-8", use_iconv = TRUE)
arrondissements <- readOGR("data/arrondissements.geojson", "arrondissements", encoding = "UTF-8", use_iconv = TRUE)
# Translate the tree sets types
levels(les_arbres@data$domanialite) <- c("Street", "Cemetery", "DAC", "School", "DASES", "Nursery", "Sports facility", "Public garden", "Beltway")
# Filter 20 arrondissements without Boulogne and Vincennes
arrondissements <- arrondissements[arrondissements@data$C_CAINSEE %/% 1000 == 75 & arrondissements@data$B_BOIS == "N" , ]
# Arrondissement descriptor reduced to its number
arrondissements@data$L_CAB <- str_replace(arrondissements@data$L_CAB, "er", "")
arrondissements@data$L_CAB <- str_replace(arrondissements@data$L_CAB, "ème", "")
# Project maps in Lambert 93 datum so that coordinates are kilometres
library(tidyverse)
les_arbres %>% 
  spTransform(CRS("+init=epsg:2154")) ->
  les_arbres_Lambert93
arrondissements %>% 
  spTransform(CRS("+init=epsg:2154")) ->
  arrondissements_Lambert93
```

Organised in sets:

- Streets, Beltway,

- Public gardens, Cemeteries,

- Schools, Nurseries, Sports facilities.

```{r, tidy=FALSE}
# Prepare the dataset
les_arbres@data %>%
  # Add coordinates in the dataset to calculate distances
  bind_cols(as.data.frame(les_arbres_Lambert93@coords)) %>% 
  # Eliminate useless columns
  select(-typeemplacement, -geo_point_2d1, -geo_point_2d2, -complementadresse) %>% 
  # Convert addresses from factors to strings for manipulation
  mutate(arrondissement=as.character(arrondissement)) %>% 
  mutate(adresse=as.character(adresse)) %>% 
  # Filter: Paris only
  dplyr::filter(str_detect(arrondissement, "^PARIS")) %>% 
  # Filter: eliminate DAC and DASES. Municipality buildings, marginal.
  dplyr::filter(domanialite!="DAC") %>% 
  dplyr::filter(domanialite!="DASES") %>% 
  # Add a column with genus-species
  mutate(spName=paste(genre, espece)) %>% 
  # Describe Arrondissement by its number only
  mutate(arrondissement=str_replace(arrondissement, "PARIS ", "")) %>% 
  mutate(arrondissement=str_replace(arrondissement, "E ARRDT", "")) %>% 
  mutate(arrondissement=str_replace(arrondissement, "ER ARRDT", "")) %>% 
  # Gather cemetery data (detailed by district) in a single set per cemetery
  mutate(adresse=str_trim(ifelse(str_detect(as.character(adresse), "^CIMETIERE "), str_split(as.character(adresse), "/", simplify=TRUE)[, 1], as.character(adresse)))) %>% 
  # Makes addresses unique when a streets covers several arrondissements or an addressed is the same for a street and a square
  mutate(adresse = paste(adresse, arrondissement, domanialite)) ->
  les_arbres_df
```

Reproducible research: code and data available.


# Measures of diversity


## Aims

We want to measure the diversity of trees in Paris areas.

- Equivalent in economics: measure the specialisation of Paris areas

Information theory is appropriate.


## Measuring Uncertainty

An experiment with several outcomes $\{r_1, \dots, r_s, \dots, r_S \}$:

- The probability to obtain $r_s$ is $p_s$.

Information function: $I(p_s)>0, p_s\in [0, 1[$, decreasing to $I(1)=0$.

- Definition: rarity is $1/p_s$.

- The logarithm of rarity is Shannon's information function.

The expectation of the information carried by an individual is Shannon's entropy:
$$\sum_s{p_s \ln {\frac{1}{p_s}}}$$

## Theil's index

Theil's  index  of specialisation is 
$$\ln{S} - \sum_s{p_s \ln {\frac{1}{p_s}}}$$

Specialisation is opposed to diversity.


## Generalized Entropy

Parametric entropy [@Tsallis1988;@Brulhart2005]

Deformed logarithm: $\ln_q x = \frac{x^{1-q} -1}{1-q}$

```{r lnq}
library("entropart")
curve(log(1/x), 0, 1, lty=1, lwd=2, ylab = expression(ln[q](1/x)))
curve(lnq(1/x, 0), 0, 1, lwd=2, lty = 2, add = TRUE)
curve(lnq(1/x, 2), 0, 1, lwd=2, lty = 3, add = TRUE)
legend("topright", 
      legend = c(expression(ln[0](1/x)), "ln(1/x)",
      expression(ln[2](1/x))),
      lty = c(2, 1, 3, 4), inset=  0.02)
```


## Formalism

Tsallis entropy is the average (deformed, of order $q$) logarithm of rarity.

The order $q$ stresses small or high probabilities.

- Entropy of order 0: the number of possible outcomes (-1)

- Entropy of order 1: Shannon ($\ln{S}-$ Theil)

- Entropy of order 2: Simpson (1-Herfindahl).


## Hill Numbers

The number of equiprobable outcomes that have the same entropy as the observed system [@Hill1973].

Exponential of entropy.

Diversity profile:

```{r}
library("SpatDiv")
q.seq <- seq(0, 2, .1)
Arrondissements <- c(8,11,16)
ADiv <- matrix(0, nrow = length(q.seq), ncol=length(Arrondissements))
for(i in 1:length(Arrondissements)) {
  les_arbres_df %>% dplyr::filter(arrondissement == Arrondissements[i]) %>% pull(spName) %>% as.AbdVector -> Abundance
  ADiv[, i] <- CommunityProfile(Diversity, Abundance, q.seq, Correction = "None")$y
}
data.frame(q=q.seq, ADiv) %>% 
  reshape::melt(id="q") %>%  
  ggplot + 
  geom_line(aes(x = q, y = value, colour = variable)) + 
  scale_colour_discrete(name = "Arrondissement", labels = c("8th", "11th", "16th"))
```

## Richness of green spaces

Diversity by type of green space: richness


```{r}
les_arbres_df %>% 
  group_by(domanialite, arrondissement, adresse) %>% 
  summarise(NumberOfTrees = n(),
            Richness = Richness(spName, Correction="None"),
            Shannon = Diversity(spName, q=1, Correction="None"),
            Simpson = Diversity(spName, q=2, Correction="None")) %>% 
  arrange(desc(Richness)) ->
  Diversite

Diversite %>% 
  ggplot(aes(x=domanialite, y=Richness)) + 
  geom_boxplot() +
  scale_y_continuous(trans = "log10") +
  labs(x="Type of green space", y="Richness")
```

## Simpson's diversity of green spaces

Diversity by type of green space: Simpson/Herfindahl


```{r}
Diversite %>% 
  ggplot(aes(x=domanialite, y=Simpson)) + 
  geom_boxplot() +
  scale_y_continuous(trans = "log10") +
  labs(x="Type of green space", y="Simpson's Diversity")
```

## Richness of arrondissements

Diversity by arrondissement: richness

```{r}
les_arbres_df %>% 
  group_by(arrondissement) %>% 
  summarise(Number = n(), 
            Richness = Richness(spName, Correction="None"),
            Shannon = Diversity(spName, q=1, Correction="None"),
            Simpson = Diversity(spName, q=2, Correction="None")) %>% 
  arrange(desc(Richness)) ->
  DiversiteArrdt

merge(arrondissements_Lambert93, DiversiteArrdt, by.x="L_CAB", by.y="arrondissement") %>% 
  spplot(zcol="Richness")
```

## Shannon's Diversity of arrondissements

Diversity by arrondissement: Shannon

```{r}
merge(arrondissements_Lambert93, DiversiteArrdt, by.x="L_CAB", by.y="arrondissement") %>% 
  spplot(zcol="Shannon")
```


## Simpson's Diversity of arrondissements

Diversity by arrondissement: Simpson

```{r}
merge(arrondissements_Lambert93, DiversiteArrdt, by.x="L_CAB", by.y="arrondissement") %>% 
  spplot(zcol="Simpson")
```



# Measures of spatial concentration

## Similar approach

We want to measure the spatial ubiquity of species.

Consider a species.

The event $r_i$ is "A tree of arrondissement $i$ belongs to the species" has probability $p_i$.

Ubiquity is the effective number of arrondissements the species is present in.

Concentration is opposed to ubiquity.


## Most abundant species

Most abundant species are plane tree, horse chestnut, scholar tree, lime tree, Norway maple, sycamore maple, European nettle tree.

```{r, warning=FALSE}
les_arbres_df %>% 
  group_by(spName) %>% 
  summarise(Number=n()) %>% 
  arrange(desc(Number))  -> AbdFreqCount
AbdFreqCount %>% 
  knitr::kable(caption="Most abundant species", longtable = TRUE, booktabs = TRUE, col.names=c("Species", "Number of Trees")) %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Spatial ubiquity of abundant species

```{r}
les_arbres_df %>% 
  group_by(spName) %>% 
  summarise(Number = n(), 
            Richness = Richness(arrondissement, Correction="None"),
            Shannon = Diversity(arrondissement, q=1, Correction="None"),
            Simpson = Diversity(arrondissement, q=2, Correction="None")) %>% 
  arrange(desc(Number)) ->
  Ubiquity


```
Spatial ubiquity of abundant species

```{r}
q.seq <- seq(0, 2, .1)
Species <- c("Platanus x hispanica", "Aesculus hippocastanum", "Sophora japonica", "Platanus occidentalis")
SpDiv <- matrix(0, nrow = length(q.seq), ncol=length(Species))
for(i in 1:length(Species)) {
  les_arbres_df %>% dplyr::filter(spName == Species[i]) %>% pull(arrondissement) %>% as.AbdVector -> Abundance
  SpDiv[, i] <- CommunityProfile(Diversity, Abundance, q.seq, Correction = "None")$y
}
data.frame(q=q.seq, SpDiv) %>% 
  reshape::melt(id="q") %>%  
  ggplot + 
  geom_line(aes(x = q, y = value, colour = variable)) + 
  scale_colour_discrete(name = "Species", labels = Species)
```

# Relative vs Absolute

## Motivation

Clarify the links between relative and absolute diversity/ubiquity.

Deal with the MAUP.

## Illustration

The diversity of parks and gardens: richness.

```{r, tidy=FALSE, message=FALSE}
# Extraction of a wmppp from an address
wmppp_adresse <- function(address, alpha=100) {
  tryCatch({
    # Prepare the dataframe
    les_arbres_df %>% 
      dplyr::filter(adresse == address) %>%
      mutate(X=coords.x1, Y=coords.x2, 
             PointType=spName, PointWeight=circonferenceencm) ->
      Points
    # At least 3 points
    if (nrow(Points) < 3) return(NA)
    
    # Build a window around the points
    if (nrow(Points) == 3) {
      # Window is a convex hull
      Window <- spatstat::convexhull.xy(Points$X, Points$Y)
    } else {
      # Window = alpha shape, i.e. concave hull
      is_validated_alpha <- FALSE
      # Use alphahull::ashape to obtain a concave hull. Parameter alpha must be small for accuracy (argument alpha) but large enough to be able to build a correct graph from the points: alpha is multiplied by 2 until success.
      while (!is_validated_alpha){
        Points %>% 
          select(x=X, y=Y) %>% 
          # Eliminate dupicates
          unique %>% 
          # Package alphahull
          alphahull::ashape(alpha=alpha) ->
          AlphaShape
        # Convert alpha shape into polygon (https://rpubs.com/geospacedman/alphasimple)
        # Make a graph with edges, library igraph
        AlphaShape_graph <- igraph::graph.edgelist(cbind(as.character(AlphaShape$edges[, "ind1"]), 
                                                         as.character(AlphaShape$edges[, "ind2"])), directed = FALSE)
        # Tests: the graph must be connected and circular. If it is not, increase alpha.
        Error <- ""
        if (AlphaShape$length == 0) {
          Error <- "No edges in alpha shape"
        } else if (!igraph::is.connected(AlphaShape_graph)) {
          Error <- "Graph not connected"
        } else if (any(igraph::degree(AlphaShape_graph) != 2)) {
          Error <- "Graph not circular"
        } else if (igraph::clusters(AlphaShape_graph)$no > 1) {
          Error <- "Graph composed of more than one circle"
        }
        if (Error == "") {
          is_validated_alpha <- TRUE
        } else {
          if (alpha > 1E8) # Unable to make a circular graph: give up.
            stop(paste(Error, address))
          else # Try to doule alpha
            alpha <- 2*alpha
        }
      }
      
      # Eliminate the first node to destroy circularity
      Cut_graph <- AlphaShape_graph - igraph::E(AlphaShape_graph)[1]
      # Find chain end points
      ends <- names(which(igraph::degree(Cut_graph) == 1))
      path <- igraph::get.shortest.paths(Cut_graph, ends[1], ends[2])[[1]]
      # This is an index into the points
      pathX <- as.numeric(igraph::V(Cut_graph)[unlist(path)]$name)
      # Join the ends to restore circularity
      pathX = c(pathX, pathX[1])
      
      # Get the points from the ashape object, make an owin. Manage reverse by tryCatch
      Window <- tryCatch(spatstat::owin(poly=list(x=AlphaShape$x[pathX, ][, 1],y=AlphaShape$x[pathX, ][, 2])),
                         error=function(e) # Error if the polygon is traversed clockwise
                           spatstat::owin(poly=list(x=AlphaShape$x[rev(pathX), ][, 1],y=AlphaShape$x[rev(pathX), ][, 2])))
    }
    
    # Create the point set
    Points %>% 
      dbmss::wmppp(window = Window, unitname = c("meter", "meters"))
  }, error = function(e) {
    # Something went wrong: return NA and print the address for future corrections.
    print(paste(e, address))
    return(NA)
  })
  
}

# convert spatstat objects to sp classes (A. Baddeley, https://stat.ethz.ch/pipermail/r-sig-geo/2009-May/005781.html)
# owin -> polygon(s)
owin2Polygons <- function(x, id="1") {
  stopifnot(spatstat::is.owin(x))
  x <- spatstat::as.polygonal(x)
  closering <- function(df) { df[c(seq(nrow(df)), 1), ] }
  pieces <- lapply(x$bdry,
                   function(p) {
                     Polygon(coords=closering(cbind(p$x,p$y)),
                             hole=spatstat.utils::is.hole.xypolygon(p))  })
  z <- Polygons(pieces, id)
  return(z)
}

# owin -> SpatialPolygons
owin2SP <- function(x) {
  stopifnot(spatstat::is.owin(x))
  y <- owin2Polygons(x)
  z <- SpatialPolygons(list(y))
  return(z)
}

# Adresse -> Polygon
adress2Polygon <- function(address) {
  ppp <- wmppp_adresse(address)
  if (any(is.na(ppp)))
    return(NA)
  else
    return(owin2Polygons(ppp$window, id=address))
}

# Make a Polygon from each address
polygons <- lapply(sort(Diversite$adresse), adress2Polygon)
# Won't work for 1 or 2 points. Eliminate NA's
polygons <- polygons[!is.na(polygons)]
# Make a SpatialPolygons object with the list of polygons
ParisTrees_sp <- SpatialPolygons(polygons)
proj4string(ParisTrees_sp) <- "+init=epsg:2154"
# Plot it
plot(ParisTrees_sp)

# Merge it with diversity
PolygonList <- sapply(ParisTrees_sp@polygons, function(x) x@ID)
Diversite_df <- as.data.frame(Diversite[Diversite$adresse %in% PolygonList, ])
row.names(Diversite_df) <- Diversite$adresse
ParisTrees_spdf <- SpatialPolygonsDataFrame (ParisTrees_sp, Diversite_df)
sp_arrondissements <- list("sp.polygons", arrondissements_Lambert93, col="green")
spplot(ParisTrees_spdf[ParisTrees_spdf@data$domanialite == "Public garden" | ParisTrees_spdf@data$domanialite == "Cemetery", ], zcol="Richness", col = NA, sp.layout = sp_arrondissements)
```

## Two types of diversity

In search for biodiversity in public gardens:

- Choose the arrondissement where public gardens are more diverse on average.

- Choose the one where they are more different from each other.

Absolute or relative diversity $\leftrightarrow$ $\alpha$ or $\beta$ diversity.


## Decomposition

Entropy(arrondissement) = entropy(within parks) + entropy(between parks)


Diversity(arrondissement) = Diversity(within parks) X Diversity(between parks)

and 

Spatial concentration(all trees) = concentration(within species) X concentration(between species)


## Within diversity of public gardens

```{r}
# Alpha
les_arbres_df %>% dplyr::filter(domanialite == "Public garden") %>% 
  group_by(arrondissement, adresse) %>% 
  summarise(NumberofTrees = n(), 
            Entropy0 = Tsallis(spName, q=0, Correction="None"),
            Entropy1 = Tsallis(spName, q=1, Correction="None"),
            Entropy2 = Tsallis(spName, q=2, Correction="None")) %>% 
  group_by(arrondissement) %>% 
  summarise(NumberOfGardens = n(), 
            alphaRichness = expq(weighted.mean(Entropy0, NumberofTrees),q=0),
            alphaShannon = expq(weighted.mean(Entropy1, NumberofTrees),q=1),
            alphaSimpson = expq(weighted.mean(Entropy2, NumberofTrees),q=2),) ->
  PGDiversite
# Plot
merge(arrondissements_Lambert93, PGDiversite, by.x="L_CAB", by.y="arrondissement") %>% 
  spplot(zcol="alphaRichness")
```


## Between diversity of public gardens

```{r}
# Plot
merge(arrondissements_Lambert93, PGDiversite, by.x="L_CAB", by.y="arrondissement") %>% 
  spplot(zcol="gammaRichness")
```


## Gamma diversity of public gardens

```{r}
# Plot
merge(arrondissements_Lambert93, PGDiversite, by.x="L_CAB", by.y="arrondissement") %>% 
  spplot(zcol="betaRichness")
```

## MAUP

Diversity/Ubiquity increases when data are aggregated: Scale effect of the MAUP.

Actually increases by between diversity : not a problem but a feature.

# Conclusions

## Key messages

Diversity $\leftrightarrow$ Specialisation

Ubiquity $\leftrightarrow$ Spatial concentration

Hill numbers unify many metrics.

Key importance of the order of diversity/ubitquity: Theil$\leftrightarrow q=1$; Ellison and Glaeser$\leftrightarrow q=2$.

Partition rather complain of the MAUP.

## Perspectives

Continuous space:

- Spatial concentration metrics

- Spatially explicit diversity metrics to develop.

```{r}
Buttes_Chaumont <- wmppp_adresse("PARC DES BUTTES CHAUMONT 19 Public garden")
plot(Buttes_Chaumont, which.marks="PointWeight")
```

## Spatially explicit diversity

Calculate the average diversity around a typical point.

```{r}
Buttes_Chaumont_accum <- Mixing(Buttes_Chaumont, q.seq=c(0,2), Individual=TRUE)
MapPlot(Buttes_Chaumont_accum, Order=0, NeighborHood=10)
```


## References

\tiny

